---
title: "Gapminder Exploratory Analysis"
author: "Delaney Callahan"
date: "09/13/2022"
output:
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(dplyr)
library(ggplot2)
library(pander)
```

## Q1: How many unique countries are represented per continent?

```{r q1, echo=TRUE}
gapminder %>% 
    group_by(continent) %>%
    summarize(count = n()) %>%
    pander(caption = "Number of Countries per Continent")
```


<br />


## Q2: Which European nation had the lowest GDP per capita in 1997?

```{r q2, echo=TRUE}
gapminder %>%
    filter(continent == "Europe" & year == 1997) %>%
    arrange(gdpPercap) %>%
    head(1) %>%
    pander(caption = "Lowest GDP in 1997")
```


<br />


## Q3: What was the average life expectancy across all continents in the 1980s?

The average life expectancy across all continents in the 1980s was `r gapminder %>% filter(year >= 1980 & year <= 1989) %>% pull(lifeExp) %>% mean() %>% round(1)`.

Unless the question was asking for life expectancy in the 1980s for each continent...

```{r q3, echo=TRUE}
gapminder %>%
    filter(year >= 1980 & year <= 1989) %>%
    group_by(continent) %>%
    summarize(mean_expectancy = mean(lifeExp)) %>%
    pander(caption = "Average Life Expectancy by Continent in the 1980s")
```


<br />


## Q4: What countries and years had a life expectancy of at least 80 years?

```{r q4, echo=TRUE}
gapminder %>%
    filter(lifeExp >= 80) %>%
    select(country, year, lifeExp) %>%
    arrange(country, year) %>%
    rename(expectancy = lifeExp) %>%
    pander(caption = "Countries that Achieved Life Expectancy over 80 and When")
```


<br />


## Q5: Select 10 rows randomly from the gapminder dataset and display them.

```{r q5, echo=TRUE}
set.seed(999)
gapminder %>%
  sample_n(size = 10, replace = FALSE) %>%
  pander(caption = "Random Sample")
```


<br />


## Q6: Except for Asia, sort the dataframe to display continent-year combinations with the highest population.

```{r q6, echo=TRUE}
gapminder %>%
    filter(continent != "Asia") %>%
    select(continent, year, pop) %>%
    arrange(desc(pop)) %>%
    head(5) %>%
    pander(caption = "Continents with Highest Population")
```


<br />


## Q7: Plot the results obtained in Q5 using `ggplot()`.

```{r q7, echo=TRUE, fig.align='center'}
set.seed(999)
random_sample <- gapminder %>%
  sample_n(size = 10, replace = FALSE)
ggplot(data = random_sample,
       aes(x = year, y = lifeExp, color = country,
           label = round(lifeExp, 1))) +
    geom_point(size = 3) +
    geom_text(nudge_y = 0.9, vjust = 0, color = "black", size = 3.5) +
    xlab("Year") +
    ylab("Life Expectancy") +
    ggtitle ("Random Sample of Life Expectancy") +
    theme_gray() +
    theme(plot.title = element_text(hjust = 0.5))
```
